---
title: "p8105_hw2_lz2950"
author: "Lehan Zou"
date: "2023-10-01"
output: github_document
---

```{r}
library(tidyverse)
library(dplyr)
library(readxl)
```

```{r}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read.csv("~/Desktop/DS/P8105_HW2/data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
```

```{r}
#read csv file
snp <- read.csv("~/Desktop/DS/P8105_HW2/data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  mutate(year = ifelse(year>=50, 1900+year, 2000+year), )|>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```

```{r}
unemployment = 
  read_csv("~/Desktop/DS/P8105_HW2/data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```
Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
```

### Problem 2
```{r}
#read csv file
Mr_trash_wheel <- read_excel("data/Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = cell_cols("A2:N586"), col_names = TRUE, trim_ws = TRUE, skip = 1) |>
  # use reasonable variable names
  janitor::clean_names() |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  mutate(homes_powered = weight_tons * 500 / 30,
         year = as.numeric(year), 
         name = "Mr Trash Wheel") 

```

```{r}
prof_trash_wheel <- read_excel("data/Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = cell_cols("A2:M108")) |>
  # use reasonable variable names
  janitor::clean_names() |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  mutate(homes_powered = weight_tons * 500 / 30,
         name = "Professor Trash Wheel") 
```

```{r}
gwy_trash_wheel <- read_excel("data/Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = cell_cols("A2:L157"), col_names = TRUE, trim_ws = TRUE, skip = 1) |>
  # use reasonable variable names
  janitor::clean_names() |>
  # omit rows that do not include dumpster-specific data
  filter(!is.na(dumpster)) |>
  mutate(homes_powered = weight_tons * (500 / 30),
         name = "Gwynnda Trash Wheel") 
```

Then we can combine those three data set.
```{r}
trash_wheel <-
  Mr_trash_wheel |> 
  full_join(prof_trash_wheel) |> 
  full_join(gwy_trash_wheel) |> 
  relocate(name)
```

The cleaned "Mr. Trash Wheel" data set has `r nrow(Mr_trash_wheel)` observations of `r ncol(Mr_trash_wheel)` variables and tells us about the trash collected by trash wheel from `r paste(Mr_trash_wheel$month[1], min(Mr_trash_wheel$year), sep = ", ")` to `r paste(Mr_trash_wheel$month[nrow(Mr_trash_wheel)], max(Mr_trash_wheel$year), sep = ", ")`. The cleaned "Professor Trash Wheel" data set contains `r nrow(prof_trash_wheel)` observations of `r ncol(prof_trash_wheel)` variables from `r paste(prof_trash_wheel$month[1], min(prof_trash_wheel$year), sep = ", ")` to `r paste(prof_trash_wheel$month[nrow(prof_trash_wheel)], max(prof_trash_wheel$year), sep = ", ")`. The cleaned "Gwynnda Trash Wheel" data set contains `r nrow(gwy_trash_wheel)` observations of `r ncol(gwy_trash_wheel)` variables from `r paste(gwy_trash_wheel$month[1], min(gwy_trash_wheel$year), sep = ", ")` to `r paste(gwy_trash_wheel$month[nrow(gwy_trash_wheel)], max(gwy_trash_wheel$year), sep = ", ")`.  
For the combined dataset, it contains `r nrow(trash_wheel)` observations of `r ncol(trash_wheel)` variables, among which the `name` variable marks the trash wheel's name (`r unique(trash_wheel$name)`). 
```{r}
professor_sum_weight = sum(pull(prof_trash_wheel, weight_tons))
gwy_ciga_butts = gwy_trash_wheel |> filter(year == 2021 & month == "July") |> pull(cigarette_butts) |> sum()
```
The total weight of trash collected by Professor Trash Wheel is `r professor_sum_weight` and the total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwy_ciga_butts`.

### Problem 3
```{r}
#read csv file
mci_baseline <-
  read_csv("data/data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    #recode sex and apoe4 to non-numeric variables
    sex = recode(sex,
                 "1" = "Male",
                 "0" = "Female"),
    apoe4 = recode(apoe4,
                   "1" = "Carrier",
                   "0" = "Non-carrier")) |>
  #remove participants who do not meet the stated inclusion criteria
  filter(age_at_onset == "."|current_age < age_at_onset )

```
For the data cleaning, first I skipped the first row which contains notes for the columns. I used recode() function to convert the numerical values of `sex` and `apoe4` to their actual meaning. Lastly, I remove participants who do not meet the stated inclusion criteria by selecting the `current_age` variable smaller than `age_at_onset` variable. After cleaning, the data set contains  `r nrow(mci_baseline)` observations of `r ncol(mci_baseline)` variables.  

From the data set, it could be concluded that `r nrow(read_csv("data/data_mci/MCI_baseline.csv", skip = 1))` participants in total were recruited, and `r nrow(mci_baseline)` participants met the criteria, `r sum(is.na(mci_baseline$age_at_onset)!=1)` develop MCI. The average baseline age is `r mean(mci_baseline$current_age)`, and `r sum(mci_baseline$sex == "Female" & mci_baseline$apoe4 == "Carrier") / sum(mci_baseline$sex == "Female") * 100`% of women in the study are APOE4 carriers.  


Next, I'm going to import the data set of longitudinally observed bionnarker values.
```{r}
#read amyloid csv file
amyloid <-
  read_csv("data/data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(baseline:time_8,
               names_prefix="time_",
               names_to="years_since_diagnosed",
               values_to="amyloid_ratio") |>
               #change baseline to 0 and convert amyloid ratio to numerical variable
               mutate(
    years_since_diagnosed =
      replace(years_since_diagnosed,years_since_diagnosed == "baseline",0),
    amyloid_ratio = as.numeric(amyloid_ratio) 
  )

amyloid
```
For the data set mci_amyloid, first I read in csv file and skip the first line, input only data entries along with variable names. Then I used `clean_names()` to make sure the names are unique and consistent. Also I renamed variable "study_id" to "id". Then I use "pivot_longer" to change the original data from horizontal to vertical with the variable (baseline:times_8) becomes "years_since_diagnosed". 
Moreover, I renamed baseline to 0 and changed "amyloid_ratio" to numeric variable.
After cleaning, There are `r nrow(amyloid)` observations of `r ncol(amyloid)` variables.

```{r}
#use anti_join() function to find unmatched cases

unmatched_baseline = anti_join(mci_baseline, amyloid, by = "id")

unmatched_amyloid = anti_join(amyloid, mci_baseline, by = "id")

```
There are `r nrow(unmatched_baseline)` for participants appear only in baseline data set and `r nrow(unmatched_amyloid)/5` appear only in amyloid data set.

```{r}
mci_combined <-
  inner_join(mci_baseline, amyloid, by= "id")
#write data to a csv file  
write_csv(mci_combined, "data/data_mci/mci_combined.csv")
```
There are `r nrow(mci_combined)/5` participants appear in both dataset, each participants have 5 rows data. So total `r nrow(mci_combined)` in the combined dataset. 
